Measuring contrast processing in the visual system using SSVEP

### 2024-04-04 review.qmd
Fixed more references

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2007] (@Hillebrand?). The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

Neurons have a limited dynamic range yet they can transmit information about visual stimuli that span many orders of magnitude. In the domain of contrast, to some extent this is accomplished at a population level - individual neurons typically implement a non-linear, sigmoidal contrast vs response (CRF) transducer [@Albrecht1982; @Tolhurst1981] and different neurons exhibit peak sensitivity (defined as the maximum slope of the function) at different contrast levels [@Carandini1994; @Carandini1998; @Busse2009]. A neuronal population will therefor span a sensitivity range greater than any individual member.

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells (@ Movshon? Levitt? Earlier... Kuffler? H+W? Alpern?) will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position. The SSVEP response to a contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

@@

In well-motivated adults, psychophysical measurements of contrast sensitivity remain the gold standard. However, it is difficult and time consuming to obtain reliable psychophysical data from infants. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [@Norcia1990; Tyler, Braddick] and the high SNR of SSVEP means that infants need only look at the screen for short periods of time.

@@

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler, Nakayama etc 1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [Tyler, @Norcia1985a; @Norcia1990].

This general approach has revealed much about the development of visual abilities in infants [@Harris1976; @Braddick1986; @Atkinson1979]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity for both achromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@Dobson1978; @Norcia1985b] with both chromatic and achromatic contrast detection reaching near-adult levels by around six months. Spatial acuity as measured by SSVEP reaches adult levels more slowly, but near-adult levels are recorded around 1 year [@Norcia1985b, @HamerVisRed1989] compared to  around 6-7 years with behavioural measures [@Ellemberg1999; @Atkinson]. At least some of this difference is likely due to the relative objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking, which require careful measurement of the infant's gaze direction. However it should be noted that other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@De-Vries-Khoe1982].

@@

The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffman; @Bach; @Farmer].

@@

*4: Clinical (Porciatti / Tsai/ Marmite/ Amblyopia / PD?)*
[@Porciatti2000; @Tsai2011]
*5: Future directions (decoding in frequency domain?, animals? BCI?)*

>>>>

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2007] @Hillebrand. The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

Neurons have a limited dynamic range yet they can transmit information about visual stimuli that span many orders of magnitude. In the domain of contrast, to some extent this is accomplished at a population level - individual neurons typically implement a non-linear, sigmoidal contrast vs response (CRF) transducer [@Albrecht1982; @Tolhurst1981] and different neurons exhibit peak sensitivity (defined as the maximum slope of the function) at different contrast levels [@Carandini1994; @Carandini1998; @Busse2009]. A neuronal population will therefore span a sensitivity range greater than any individual member.

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the LGN [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position. The SSVEP response to a contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

@@

In well-motivated adults, psychophysical measurements of contrast sensitivity remain the gold standard. However, it is difficult and time consuming to obtain reliable psychophysical data from infants. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [@Norcia1990; @Tyler1979; @Braddick1986] and the high SNR of SSVEP means that infants need only look at the screen for short periods of time.

@@

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979, @Norcia1985a; @Norcia1990].

This general approach has revealed much about the development of visual abilities in infants [@Harris1976; @Braddick1986; @Atkinson1979]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity for both achromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@Dobson1978; @Norcia1985b] with both chromatic and achromatic contrast detection reaching near-adult levels by around six months. Spatial acuity as measured by SSVEP reaches adult levels more slowly, but near-adult levels are recorded around 1 year [@Norcia1985b; @Hamer1989] compared to  around 6-7 years with behavioural measures [@Ellemberg1999; @Atkinson]. At least some of this difference is likely due to the relative objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking, which require careful measurement of the infant's gaze direction. However it should be noted that other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@De-Vries-Khoe1982].

@@

The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffman2017; @Bach2008; @Bach2020].

@@

*4: Clinical (Porciatti / Tsai/ Marmite/ Amblyopia / PD?)*

*5: Future directions (decoding in frequency domain?, animals? BCI?)*
